\documentclass[a4paper,11pt,draft]{memoir}

\semiisopage
\checkandfixthelayout
\chapterstyle{tandh}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

\usepackage{lmodern}
\usepackage{roboto}
\usepackage{mathptmx}
\usepackage{tgtermes}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{stmaryrd}

\usepackage{mathtools}
\usepackage{amsthm}
\usepackage{thmtools}
\declaretheorem{theorem}
\declaretheorem{lemma}
\declaretheorem{corollary}
\declaretheorem[style=definition,qed=$\blacktriangle$]{example}
\declaretheorem[style=definition,qed=$\blacktriangle$]{definition}

\usepackage{enumitem}
\setlist[enumerate,1]{label=\alph*)}
\newlist{prfcases}{enumerate}{1}
\setlist[prfcases,1]{label={\itshape Case \arabic*.},align=left,leftmargin=\parindent,itemindent=*}

\usepackage{algorithm}
\usepackage[final]{listings}
\lstset{
	numbers=left,
	firstnumber=1,
	stepnumber=5,
	numberfirstline,
	basicstyle=\ttfamily\small,
	keywordstyle=\ttfamily\small\bfseries,
	numberstyle=\footnotesize,
	columns=flexible
}

\usepackage{tikz}
\usetikzlibrary{shapes.geometric,matrix,decorations.pathreplacing}
\tikzset{itermtree/.style={level distance=6mm,sibling distance=10mm,%
	edge from parent/.style={draw,solid},every node/.style={solid}}}
\tikzset{pure/.style={draw,rectangle,inner sep=4pt,minimum size=5mm}}
\tikzset{term/.style={draw,circle,inner sep=1pt,minimum size=5mm}}
\tikzset{ap/.style={fill,diamond,inner sep=0pt,minimum size=6pt}}
\tikzset{subtrm/.style={draw,isosceles triangle,isosceles triangle apex angle=55,%
	shape border rotate=90,minimum height=10mm,anchor=north,child anchor=north}}
\tikzset{subtrmf/.style={subtrm,level distance=10mm,sibling distance=16.67mm}}
\tikzset{subtrmn/.style={subtrm,level distance=4mm,sibling distance=6.67mm}}
\tikzset{abbrv/.style={dashed}}

\usepackage[english]{babel}
\hyphenation{Isa-belle}

\usepackage[
	style=numeric,
	doi=false,
	urldate=long,
]{biblatex}
\DeclareFieldFormat{url}{\newline\url{#1}}
\addbibresource{bibliography.bib}

\usepackage[final,hidelinks]{hyperref}


\newcommand{\ldb}{\llbracket}
\newcommand{\rdb}{\rrbracket}
\newcommand{\todo}{\fbox{To do.}}

\newcommand{\oftype}{\mathrel{::}}
\newcommand{\funT}{\Rightarrow}
\newcommand{\abs}[2]{\lambda #1.\>#2}
\newcommand{\Imp}{\Longrightarrow}
\newcommand{\imp}{\longrightarrow}
\newcommand{\Iff}{\Longleftrightarrow}
\newcommand{\All}[2]{\bigwedge #1.\>#2}
\newcommand{\all}[2]{\forall #1.\>#2}
\newcommand{\tvar}[1]{{'\!\textit{#1}}}
\newcommand{\svar}[1]{\textit{?#1}}
\newcommand{\set}[2]{\{#1\mid #2\}}

\DeclareMathOperator{\pure}{\mathit{pure}}
\newcommand{\ap}{\diamond}
\DeclareMathOperator{\sterm}{\mathsf{term}}
\DeclareMathOperator{\spure}{\mathsf{pure}}
\DeclareMathOperator{\sivar}{\mathsf{var}}
\newcommand{\siabs}[2]{\mathsf{abs}\;#1.\>#2}
\newcommand{\sap}{\mathbin{\mathsf{`ap`}}}
\newcommand{\sapp}{\>}
\newcommand{\sabs}[2]{{\boldsymbol\lambda} #1.\>#2}
\newcommand{\tabs}[2]{{\boldsymbol\lambda}^\ast #1.\>#2}
\newcommand{\termeq}{=_{\alpha\beta\eta}}
\newcommand{\unlift}[1]{\left\downarrow #1\right.}
\DeclareMathOperator{\vars}{var}
\newcommand{\varseq}{\overrightarrow{\vars}}
\DeclareMathOperator{\abseq}{abs}

\newenvironment{isabelle}{\list{}{}\item\relax}{\endlist}
\newcommand{\iindent}{\-\hspace{1em}}


\begin{document}

\begin{titlingpage}
\calccentering{\unitlength}
\begin{adjustwidth*}{\unitlength}{-\unitlength}
\sffamily\large\raggedleft
{\huge\bfseries Applicative Functors in Isabelle/HOL} \\
\vspace{7mm}
{\Large Joshua Schneider} \\
\vspace{2cm}
Bachelor Thesis \\
(11-947-173) \\
\vspace{7mm}
\makeatletter\@date\makeatother

\vspace{\fill}
Advisors: \\
Prof.\ Dr.\ David Basin \\
Dr.\ Andreas Lochbihler \\
\vspace{7mm}
Department of Computer Science, ETH Zürich
\end{adjustwidth*}
\end{titlingpage}

\pagestyle{plain}
\begin{abstract}
Applicative functors can be found in different contexts.
In functional programming, they abstract the addition of effects to types.
But also rather mathematical constructs like sets and infinite containers
are examples.
The key property of applicative functors is that they lift functions
to effectful computations.

We present a proof method based on applicative lifting, implemented for the
Isa\-belle/HOL proof assistant.
Hinze showed that properties of functions are often preserved when lifted.
While functors are represented naturally in the types of HOL, it is not
directly possible to reason based on this insight.
We propose an algorithm which computes a certain normal form of applicative
expressions, together with a checked proof of equivalence.
We explain formally how this normal form relates base and lifted equations.
Because not all equations can be handled this way, we apply abstraction
algorithms, as known from combinatory logic, to applicative expressions.
Depending on the functor, different combinators can be lifted, which the
increases the flexibility of our proof method.
We choose a fixed collection of combinators, which we consider to cover
relevant cases.
Scope and implementation of each case are discussed.
We provide the combined development as a ready-to-use package for Isabelle, and
demonstrate its usability on algebraic structures lifted to streams and
infinite trees.
\end{abstract}

\cleartorecto
\tableofcontents*

\pagestyle{plain}
\input{introduction}
\input{design}
\input{normal_form}
\input{combinators}
\input{examples}
\input{conclusion}

\cleartorecto
\emergencystretch 1em
\printbibliography[heading=bibintoc]

\end{document}
